---
title: "Adelie Penguins and Mediation Effect"
Author: M. Hyde
date: "2024-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**This code uses the penguins dataset and regresses `bill_length_mm` on `flipper_length_mm` and `body_mass_g` to perform the Sobel test for mediation effect.**

# Importing data

```{r}
# Install and load necessary packages
if (!require("palmerpenguins")) install.packages("palmerpenguins")
library(palmerpenguins)

# Load the penguins dataset
data("penguins")
str(penguins)

# Check for missing values
sum(is.na(penguins))
```

# Remove rows with missing values
```{r}
# Remove rows with missing values
penguins <- na.omit(penguins)

head(penguins)
```
# Select only the species "Adelie"
```{r}
penguins <- subset(penguins, species == "Adelie")

```

# Problem 1

Given the provided dataset "penguins", consider the model regressing `bill_length_mm` on `flipper_length_mm`, and `body_mass_g`. Conduct a Sobel test to see whether there is a mediation effect. You need to consider the three relevant models:

1. Model 1: \(bill\_length_i = \beta_{11} + \beta_{12}flipper\_length_i + \epsilon_i\)
2. Model 2: \(body\_mass_i = \beta_{21} + \beta_{22}flipper\_length_i + \epsilon_i\)
3. Model 3: \(bill\_length_i = \beta_{31} + \beta_{32}body\_mass_i + \beta_{33}flipper\_length_i + \epsilon_i\)

Suppose the variance of \(\epsilon_i\) is known with \(\sigma^2 = 1\), for all three models. Use the level of significance \(\alpha\)=0.05.

**(1) What is the null hypothesis of the Sobel test?**

The null hypothesis of the Sobel test is that there is no significant indirect effect or mediation effect.

\(H_0\): \(\beta_{12}=\beta_{33} = 0\)

This hypothesis implies that the mediator variable `body_mass_g` does not significantly mediate the relationship between the independent variable `flipper_length_mm` and the dependent variable `bill_length_mm`.

**(2) Regress the three models and find the estimators for (β11, β12), (β21, β22), and (β31, β32, β33).**

```{r}
# Model 1: bill_length_i = β11 + β12 * flipper_length_i + εi
model1 <- lm(bill_length_mm ~ flipper_length_mm, data = penguins)
beta_11_12 <- coef(model1)
print(beta_11_12)

# Model 2: body_mass_i = β21 + β22 * flipper_length_i + εi
model2 <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
beta_21_22 <- coef(model2)
print(beta_21_22)

# Model 3: bill_length_i = β31 + β32 * body_mass_i + β33 * flipper_length_i + εi
model3 <- lm(bill_length_mm ~ body_mass_g + flipper_length_mm, data = penguins)
beta_31_32_33 <- coef(model3)
print(beta_31_32_33)
```

**(3) Find the variance of \(\hat{\beta}_{22}\) and \(\hat{\beta}_{32}\).**

```{r}
# Extract covariance matrix
cov_matrix_model2 <- vcov(model2)
cov_matrix_model3 <- vcov(model3)

# Variance of beta22 in Model 2
var_beta22_model2 <- cov_matrix_model2["flipper_length_mm", "flipper_length_mm"]
cat("Variance of beta22 in Model 2:", var_beta22_model2, "\n")

# Variance of beta32 in Model 3
var_beta32_model3 <- cov_matrix_model3["body_mass_g", "body_mass_g"]
cat("Variance of beta32 in Model 3:", var_beta32_model3, "\n")
```

**(4) Find the `z` test statistic.**

The z test formula

\[ Z = \frac{\hat{B}_{12} - \hat{B}_{33}}{\sqrt{\hat{B}_{22}^2 \cdot \text{Var}(\hat{B}_{32}) + \hat{B}_{32}^2 \cdot \text{Var}(\hat{B}_{22})}} \]

```{r}
# Coefficient estimates
beta12 <- coef(model1)["flipper_length_mm"]
beta22 <- coef(model2)["flipper_length_mm"]
beta33 <- coef(model3)["flipper_length_mm"]
beta32 <- coef(model3)["body_mass_g"]
var_beta32 <- cov_matrix_model3["body_mass_g", "body_mass_g"]
var_beta22 <- cov_matrix_model2["flipper_length_mm", "flipper_length_mm"]

# Calculate the Sobel test statistic
z_test_statistic <- (beta12 - beta33) / sqrt(beta22^2 * var_beta32 + beta32^2 * var_beta22)
print(z_test_statistic)
```

**(5) Find the `p − value` and the conclusion of the test**

Using the formula \( p = 2(1 - \Phi(|z|)) \):

```{r}
# Using results from z_test_statistic calculated
p_value_sobel_formula <- 2 * (1 - pnorm(abs(z_test_statistic)))
print(p_value_sobel_formula)
```

The p-value is less than 0.05. Hence, we reject the Null Hypothesis. There is a significant mediation effect.


